State of the Practice for Lattice Boltzmann Method Software

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits from LanguageTool corrections

\affiliation[CAS]{organization={McMaster University, Computing and Software
Department},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ME]{organization={McMaster University, Mechanical Engineering},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

@@

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of 20 questions). After measuring we ranked the software using
    the Analytic Hierarchy Process (AHP). Our ranking was roughly consistent
    with GitHub stars ranking, suggesting at least a partial correlation
    between the use of best practices and popularity. We find the state of the
    practice to be healthy, with 67\% of the measured packages ranking in the
    top five for at least one quality, the majority of LBM generated artifacts
    corresponding to general recommendations from research software developers,
    common use of version control (67\% of packages) and the adoption of a
    quasi-agile development process.  Areas of best practice to potentially
    improve include adoption of continuous integration, API documentation and
    enforcement of programming style guides.  We interviewed four developers to
    gain insight into their current pain points. Identified challenges include
    lack of development time, lack of funding, and difficulty with ensuring
    correctness.  Developers are addressing these pain points by designing for
    change, circumventing the oracle problem and prioritizing documentation and
    usability. For future improvements we suggest the following: employing
    linters, conducting rigorous peer reviews, writing and submitting more
    papers on software, growing the number of contributors by following current
    recommendations for open source projects, and augmenting the theory manuals
    to include more requirements relevant information.

@@

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behaviour of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two-step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum are
transferred among particles that collide \citep{bao2011lattice}. As an example
of an important application of LBM, Figure~\ref{Fig_coarctation} shows the flow
in a defective aorta pre and post-medical intervention. LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of safety and mission-critical applications of LBM includes the
following: designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater
flow \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the
cardiovascular system \citep{SadeghiEtAl2022b, SadeghiEtAl2022,
SadeghiEtAl2020}.  As these examples illustrate, engineers and scientists use
LBM for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety. Given the critical applications of LBM,
users of the libraries are concerned with software qualities like reliability,
robustness, reproducibility, performance, correctness and verifiability.  Since
their time is valuable, developers of LBM libraries have additional quality
concerns, including maintainability, reusability, and understandability.  With
the quality considerations in mind, and the potentially overwhelming number of
choices for existing LBM libraries, the time is right to assess the state of the
practice. Therefore, we analyze the current state of LBM
software development and compare it to the development practices employed for
research software in general.  We want to highlight success stories that LBM
developers can share amongst their community, and with the broader research
software community, while at the same time looking for areas for potential
future improvement.

We use ten research questions to structure and focus our discussion. Our data
collection methodology \citep{SmithEtAl2021, SmithAndMichalski2022} inspires the
questions.  For each question below, we point to the section that contains our
answer.  We start with identifying the examples of LBM software where we have
access to the source code:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  By \emph{best practices}, we
mean methods, techniques and tools that are generally believed to improve
software development, like testing and documentation.  We will categorize our
best practices around software qualities, as we will discuss in
Section~\ref{empiricalmeasures}.  At this point in the process, to remove
potential user/developer bias, we will base our assessment only on publicly
available artifacts, where artifacts are the documents, scripts and code that we
find in a project's public repository. Example artifacts include requirements,
specifications, user manuals, unit test cases, system tests, usability tests,
build scripts, API (Application Programming Interface) documentation, READMEs,
license documents, process documents, and code. Following best practices does
not guarantee popularity, so we will also compare our ranking to how the user
community ranks the identified projects.

@@

To understand the state of the practice, we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how LBM software compares to other research software.  For cases
where data is available, we can compare to other specific domains of research
software, like medical imaging \citep{Dong2021} and ocean modelling
\citep{JungEtAl2022}.  However, most of the literature on the state of the
practice does not focus on specific domains, but rather on all research software
\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
not distinguish an application domain as \emph{research software in general}. By
comparing LBM software to research software in general we can determine how LBM
follows the trends from other developer communities and how it is different.

\begin{enumerate}
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used?
	(Section~\ref{Sec_CompareMethodologies})
\end{enumerate}

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns and how they deal with them; we need to understand their pain
points. We wish to know what practices are used by the top LBM projects, so that
others can potentially emulate these practices. We also wish to identify new
practices that LBM developers can adopt to improve their software in the future.
The question below cover these points:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers what specific best practices are taken to address the pain points
	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practice could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}).
	(Section~\ref{Sec_Recommendations})
\end{enumerate}

@@

Compared with our previous methodology (as used in \citet{SmithEtAl2016} for
instance), measuring the application of best practices now also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc.  We rank the software with the
Analytic Hierarchy Process (AHP), using the quantitative data from the grading
template. After this, as another addition to our previous methodology, we
interview some of the development teams to understand the status of their
development processes. We next answer the research questions using the collected
quantitative and qualitative data (Sections~\ref{methodology}
to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
(Section~\ref{threats}) and the final conclusions and recommendations
(Section~\ref{conclusion}).

@@

To ensure our assessment of best practices is organized and complete, we centre
our measurements around software qualities. For completeness, we use the list of
software engineering qualities from \citet[p.\ 17--33]{GhezziEtAl2003},
excluding interoperability, productivity and timeliness. We excluded
interoperability because our focus is on LBM software alone, not its interaction
with other software. Our exclusion of the other two qualities is because they
are process qualities, and our focus is on product qualities. We added
installability to our list since potential users will ignore software they
cannot install.  Since the scientific method depends on reproducibility, we also
added this quality. Below we provide information on each quality, including
details on how they are measured.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case, the effort includes the following: time spent finding and
	understanding the installation instructions, the person-time and resources
	spent performing the installation procedure, and the absence or ease of
	overcoming system compatibility issues. An installability score improves if
	the software has means to validate the installation and instructions for
	uninstallation.

@@

	\item[Performance] Performance is measured by the degree to which a system
	or component accomplishes its designated functions within given constraints,
	such as speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing
	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
	In this state of the practice assessment performance is not directly
	quantitatively measured. Instead we check the documentation for each
	software package looking for information alluding to consideration of
	performance, such as testing results or parallelization instructions.

@@

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
	large number of easily reusable components. Therefore, we look for high
	modularization, which we define as the presence of small components with
	well defined interfaces. For this state of the practice assessment, we
	assume that a good measure of reusability correlates with a high number of
	code files, and the availability of API documentation.

@@

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all of the steps of a software development process, and its
	current status, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}. In this
	state of the practice assessment a good measure of visibility and
	transparency correlates with a well-defined development process,
	documentation of the development process and environment, and software
	version release notes.

@@

For the third item in the above filter, we categorized software packages as
`alive' if documentation updates happened within the last 18 months. Packages
were categorized as `dead' if their last update were more than 18 month ago.

@@

We designed the questions to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. We categorize them by quality, with an extra
category for summary information (Figure~\ref{measurement_template_image}). The
summary section provides general data, such as the software name, number of
developers, etc. \citet{gewaltig2012quality} provide us with our software
categories. Public means software available to all, while private means software
available only to a specific group.  The concept category defines software
written to demonstrate algorithms or concepts. The three categories of
development models are open source, where source code is freely available under
an open-source license; free-ware, where a binary or executable is provided for
free; and commercial, where the user must pay.

@@

We requested interviews with developers for all 24 packages, except Munubi,
since we added Munubi after our ethics approval expired. We found the developers
from project websites, code repositories, publications, and biographic
information on institution webpages. Four developers agreed to participate in
our study. We met with them online, using either Zoom or Teams to facilitate
recording and automatic transcription. Each meeting lasted about an hour. The
McMaster University Research Ethics Board approved the interview process under
application number
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions and assumptions that
differ between family members. Associated with each variability are parameters
of variation, which summarize the possible values for that variability, along
with their potential binding time. The binding time is when the value of a
variability is set. Variabilities are set at specification time, build time
(when the program compiles) or run time (when the code executes). For the
current assessment, the packages all have the commonality of using LBM
techniques to simulate the motion of fluids. Variabilities for LBM packages
include the following: the dimension of problems they simulate, whether or not
(and how) they support parallel computation, whether or not compressibility can
be considered, whether reflexive boundaries can be employed, whether the
simulation handles multi-fluids, whether the simulations handles turbulence, and
whether complex geometries can be simulated.

@@

\begin{table}[ht!]
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\toprule
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\midrule
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			\esp{} & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\bottomrule
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
			Complex Geometries (Yes or No), OS for Operating System (Windows
			(W), macOS (M), Linux (L))) (Sorted Alphabetically)} \label{tbl_features}
	\end{center}
\end{table}

@@

All but one of the software packages (LatBo.jl) have automated at least some of
the installation process. Most of these packages, such as waLBerla and
SunlightLB, use Make to automate the installation, and a few of them, like
lbmpy, use custom scripts.

@@

Figure~\ref{Fig_Installability} shows the installability ranking of the software
packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL
MESO, Palabos) tend to have one set of linear installation instructions written
as if the person doing the installation has none of the dependencies installed.
The instructions often list compatible operating system versions and include
instructions for installing dependencies. The top-ranked packages often
incorporate some automation of the installation process. The number of
dependencies a package has does not correlate with a higher score. The ability
to validate the installation process, say through tutorials or test cases, leads
to a higher score. Apparently active work on a project improves installability
since the top seven ranked packages are all classified as alive.

@@

Seventeen of the software packages explicitly reference domain theory, but a
full and rigorous requirements specification, in the software engineering sense
\citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991}, is absent from all of
the projects. Software packages that include a subset of a requirements
specification, such as DL\_MESO (LBE), keep the information brief and include it
within other documents. When present, theory related information is generally
found within a user manual, on a webpage, or in a cited publication. In the
latter case significant time may be needed to track down the information.

@@

Figure~\ref{Fig_CorrectnessVerifiability} shows the ranking for surface
correctness and verifiability. Software packages with a higher score tend to
have theory documentation. They also explicitly use at least one document
generation tool that helps with verification, thus building confidence in
correctness. The top ranked software packages all include an easy to follow
getting started tutorial, and most of these include expected output. The two top
ranked packages, Ludwig and Musubi, indicate the use of unit testing in their
documentation. They also incorporated CI in the development
process. Furthermore, eight of the top 10 ranked packages are noted as being
alive.

@@

Figure~\ref{Fig_Reliability} shows the surface reliability ranking of the
software packages. Software packages with a high score either did not break
during installation, or the broken installation was recoverable. All of the top
five ranked packages have tutorials. The package pyLBM broke during tutorial
testing, but a descriptive error message helped in recovery. Nine of the top 10
ranked packages are noted as being alive.

>>>>

\affiliation[CAS]{organization={McMaster University, Computing and Software
Department}, %Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

@@

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of 20 questions). After measuring, we ranked the software using
    the Analytic Hierarchy Process (AHP). Our ranking was roughly consistent
    with GitHub stars ranking, suggesting at least a partial correlation
    between the use of best practices and popularity. We find the state of the
    practice to be healthy, with 67\% of the measured packages ranking in the
    top five for at least one quality, the majority of LBM generated artifacts
    corresponding to general recommendations from research software developers,
    common use of version control (67\% of packages) and the adoption of a
    quasi-agile development process.  Areas of best practice to potentially
    improve include adoption of continuous integration, API documentation and
    enforcement of programming style guides.  We interviewed four developers to
    gain insight into their current pain points. Identified challenges include
    lack of development time, lack of funding, and difficulty with ensuring
    correctness.  Developers are addressing these pain points by designing for
    change, circumventing the oracle problem and prioritizing documentation and
    usability. For future improvements we suggest the following: employing
    linters, conducting rigorous peer reviews, writing and submitting more
    papers on software, growing the number of contributors by following current
    recommendations for open source projects, and augmenting the theory manuals
    to include more requirements relevant information.

@@

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM solvers consider the behaviour of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. These solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two-step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where particles transfer energy
and momentum when they collide \citep{bao2011lattice}. As an example of an
important application of LBM, Figure~\ref{Fig_coarctation} shows the flow in a
defective aorta pre- and post-medical intervention. LBM has several advantages
over conventional CFD methods, including a simple calculation procedure,
improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of safety and mission-critical applications of LBM includes the
following: designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater
flow \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the
cardiovascular system \citep{SadeghiEtAl2022b, SadeghiEtAl2022,
SadeghiEtAl2020}.  As these examples illustrate, engineers and scientists use
LBM for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety. Given the critical applications of LBM,
users worry about with software qualities like reliability,
robustness, reproducibility, performance, correctness, and verifiability.  Since
their time is valuable, developers of LBM libraries have additional quality
concerns, including maintainability, reusability, and understandability.  With
the quality considerations in mind, and the potentially overwhelming number of
choices for existing LBM libraries, the time is right to assess the state of the
practice. Therefore, we analyze the current state of LBM
software development and compare it to the development practices employed for
research software in general.  We want to highlight success stories that LBM
developers can share amongst their community, and with the broader research
software community, while at the same time looking for areas for potential
future improvement.

We use ten research questions to structure and focus our discussion. Our
methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by the
questions. For each question below, we point to the section that contains our
answer.  We start with identifying the examples of LBM software where we have
access to the source code:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  By \emph{best practices}, we
mean methods, techniques, and tools that are generally believed to improve
software development, like testing and documentation.  We will categorize our
best practices around software qualities, as we will discuss in
Section~\ref{empiricalmeasures}.  At this point in the process, to remove
potential user/developer bias, we will base our assessment only on publicly
available artifacts, where artifacts are the documents, scripts, and code that we
find in a project's public repository. Example artifacts include requirements,
specifications, user manuals, unit test cases, system tests, usability tests,
build scripts, API (Application Programming Interface) documentation, READMEs,
license documents, process documents, and code. Following best practices does
not guarantee popularity, so we will also compare our ranking to how the user
community ranks the identified projects.

@@

To understand the state of the practice, we wish to learn the frequency with
which different artifacts appear, the types of development tools used, and the
methodologies used for software development.  With this data, we can ask
questions about how LBM software compares to other research software.  For cases
where data is available, we can compare to other specific domains of research
software, like medical imaging \citep{Dong2021} and ocean modelling
\citep{JungEtAl2022}.  However, most of the literature on the state of the
practice does not focus on specific domains, but rather on all research software
\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
not distinguish an application domain as \emph{research software in general}. By
comparing LBM software to research software in general we can determine how LBM
follows the trends from other developer communities and how it is different.

\begin{enumerate}
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes, and methodologies used?
	(Section~\ref{Sec_CompareMethodologies})
\end{enumerate}

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns and how they deal with them; we need to understand their pain
points. We wish to know what practices top LBM projects use, so that
others can potentially emulate these practices. Part of our goal is to identify new
practices that LBM developers can adopt to improve their software in the future.
The question below cover these points:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers what specific best practices address the pain points
	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practices could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}.
	(Section~\ref{Sec_Recommendations})
\end{enumerate}

@@

Compared with our previous methodology (as used in \citet{SmithEtAl2016} for
instance), measuring the application of best practices now also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc.  We rank the software with the
Analytic Hierarchy Process (AHP), using the quantitative data from the grading
template. After this, as another addition to our previous methodology, we
interview some development teams to understand the status of their
development processes. We next answer the research questions using the collected
quantitative and qualitative data (Sections~\ref{methodology}
to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
(Section~\ref{threats}) and the final conclusions and recommendations
(Section~\ref{conclusion}).

@@

To ensure our assessment of best practices is organized and complete, we centre
our measurements around software qualities. For completeness, we use the list of
software engineering qualities from \citet[p.\ 17--33]{GhezziEtAl2003},
excluding interoperability, productivity, and timeliness. We excluded
interoperability because our focus is on LBM software alone, not its interaction
with other software. Our exclusion of the other two qualities is because they
are process qualities, and our focus is on product qualities. We added
installability to our list since potential users will ignore software they
cannot install.  Since the scientific method depends on reproducibility, we also
added this quality. Below we provide information on each quality, including
details on how they are measured.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation, or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case, the effort includes the following: time spent finding and
	understanding the installation instructions, the person-time and resources
	spent performing the installation procedure, and the absence or ease of
	overcoming system compatibility issues. An installability score improves if
	the software has means to validate the installation and instructions for
	uninstallation.

@@

	\item[Performance] Performance is measured by the degree to which a system
	or component accomplishes its designated functions within given constraints,
	such as speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing
	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
	In this state of the practice assessment performance is not directly
	quantitatively measured. Instead, we check the documentation for each
	software package looking for information alluding to consideration of
	performance, such as testing results or parallelization instructions.

@@

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good indicator of reusability is many
	easily reusable components. Therefore, we look for high modularization,
	which we define as the presence of small components with well-defined
	interfaces. For this state of the practice assessment, we assume that a good
	measure of reusability correlates with a high number of code files, and the
	availability of API documentation.

@@

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all the steps of a software development process, and its
	current status, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}. In this
	state of the practice assessment a good measure of visibility and
	transparency correlates with a well-defined development process,
	documentation of the development process and environment, and software
	version release notes.

@@

For the third item in the above filter, we categorized software packages as
`alive' if documentation updates happened within the last 18 months. Packages
were categorized as `dead' if their last update were more than 18 months ago.

@@

We designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We categorize them by quality, with an extra
category for summary information (Figure~\ref{measurement_template_image}). The
summary section provides general data, such as the software name, number of
developers, etc. \citet{gewaltig2012quality} provide us with our software
categories. Public means software available to all, while private means software
available only to a specific group.  The concept category defines software
written to demonstrate algorithms or concepts. The three categories of
development models are open source, where source code is freely available under
an open-source license; free-ware, where a binary or executable is provided for
free; and commercial, where the user must pay.

@@

We requested interviews with developers for all 24 packages, except Musubi,
since we added Musubi after our ethics approval expired. We found the developers
from project websites, code repositories, publications, and biographic
information on institution webpages. Four developers agreed to participate in
our study. We met with them online, using either Zoom or Teams to facilitate
recording and automatic transcription. Each meeting lasted about an hour. The
McMaster University Research Ethics Board approved the interview process under
application number
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities, and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions, and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions, and assumptions that
differ between family members. Associated with each variability are parameters
of variation, which summarize the possible values for that variability, along
with their potential binding time. The binding time is when the value of a
variability is set. Variabilities are set at specification time, build time
(when the program compiles) or run time (when the code executes). For the
current assessment, the packages all have the commonality of using LBM
techniques to simulate the motion of fluids. Variabilities for LBM packages
include the following: the dimension of problems they simulate, whether 
(and how) they support parallel computation, whether compressibility can
be considered, whether reflexive boundaries can be employed, whether the
simulation handles multi-fluids, whether the simulation handles turbulence, and
whether complex geometries can be simulated.

@@

\begin{table}[ht!]
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\toprule
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\midrule
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			\esp{} & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & — & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & — & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & — & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\bottomrule
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
			Complex Geometries (Yes or No), OS for Operating System (Windows
			(W), macOS (M), Linux (L))) (Sorted Alphabetically)} \label{tbl_features}
	\end{center}
\end{table}

@@

All but one of the software packages (LatBo.jl) have automated at least part of
the installation process. Most of these packages, such as waLBerla and
SunlightLB, use Make to automate the installation, and a few of them, like
lbmpy, use custom scripts.

@@

Figure~\ref{Fig_Installability} shows the installability ranking of the software
packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
DL\_MESO, Palabos) tend to have one set of linear installation instructions
written as if the person doing the installation has none of the dependencies
installed. The instructions often list compatible operating system versions and
include instructions for installing dependencies. The top-ranked packages often
incorporate some automation of the installation process. The number of
dependencies a package has does not correlate with a higher score. The ability
to validate the installation process, say through tutorials or test cases, leads
to a higher score. Apparently active work on a project improves installability
since the top seven ranked packages are all classified as alive.

@@

Seventeen software packages explicitly reference domain theory, but a complete
and rigorous requirements specification, in the software engineering sense
\citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991},  is absent from all
projects. Software packages that include a subset of a requirements
specification, such as DL\_MESO (LBE), keep the information brief and include it
within other documents. When present, theory related information is generally
found within a user manual, on a webpage, or in a cited publication. In the
latter case tracking down the knowledge may take a significant time.

@@

Figure~\ref{Fig_CorrectnessVerifiability} shows the ranking for surface
correctness and verifiability. Software packages with a higher score tend to
have theory documentation. They also explicitly use at least one document
generation tool that helps with verification, thus building confidence in
correctness. The top ranked software packages all include getting started
tutorials, and most of these include expected output. The two top ranked
packages, Ludwig and Musubi, indicate the use of unit testing in their
documentation. They also incorporated CI in the development process.
Furthermore, eight of the top 10 ranked packages are noted as being alive.

@@

Figure~\ref{Fig_Reliability} shows the surface reliability ranking of the
software packages. Software packages with a high score either did not break
during installation, or the broken installation was recoverable. All the top
five ranked packages have tutorials. The package pyLBM broke during tutorial
testing, but a descriptive error message helped in recovery. Nine of the top 10
ranked packages are noted as being alive.
